Visualizing Functional Data

### 2016-11-03 main.tex
moved section text into proper sections

High dimensional data is catch-all term, typically used to describe data that's highly
multivariate or data that's structurally complex, or even data that simply has
a large number of observations. The lack of specificity of the phrase
`high-demensional` can make it difficult to parse out which high-dimensional
analysis and visaualization technique is suited to the research question at
hand. Datasets inherently have two components, the variables and the
observations of those variables. Functional data analysis provides a language
for formalizing the structure of the data in such a way that it clarifies the
possible interactions in the data.

%%paragraph on functional data analysis

Observations can often be reduced to one of the following forms:

\begin{tabular}
   & univarate   & bivariate                & multivariate \\
0D & v_i         & v0_i, v1_i               & v0_i, \dots, vn_i \\
1D & v_i(x_i)     & v0_i(x_i), v1_i(x_i)    & v0_i(x_i), \dots, vn_i(x_i)\\
ND & v_i(x0_i,\dots,xn_i)     & v0_i(x0_i,\dots,xn_i), v1_i(x0_i,\dots,xn_i)    & v0_i(x0_i,\dots,xn_i), \dots, vn_i(x0_i,\dots,xn_i)\\
\end{tabular}

The variables $v0,\dots, vn$ can also potentially be dependent on each other, adding in a third level of interactions that a visualization may want to caputure.

>>>>

High dimensional data is catch-all term, typically used to describe data that's highly
multivariate or data that's structurally complex, or even data that simply has
a large number of observations. The lack of specificity of the phrase
`high-demensional` can make it difficult to parse out which high-dimensional
analysis and visaualization technique is suited to the research question at
hand. Datasets inherently have two components, the variables and the
observations of those variables. These oberservations are sometimes dependent
on information rich parameters such as time and space, and often it is as
important to visualize these underlying parametric dependencys as the
relationships between the variables in given observations. Functional data
analysis provides a language for formalizing the structure of the data in such a way that it clarifies the
possible interactions in the data.

%%paragraph on functional data analysis
The underlying assumption of functional data analysis is that while sometimes
samples are discrete, often they are drawn from a continuos sample of
obersvations \cite{ramsey, hook, meuller}. Independent observations
by definition rely on zero paramters, whereas dependent observations can be
paramterized on one dimension (such as as time), or multiple (such as time,
latitude, longitude, altitude). %%throw some math here to transition to my table

\begin{table}
\begin{tabular}
   & univarate   & bivariate                & multivariate \\
0 parameter & RV_i         & v0_i, v1_i               & RV0_i, \dots, RVn_i \\
1 paramter & RV_i(x_i)     & RV0_i(x_i), RV1_i(x_i)    & RV0_i(x_i), \dots, RVn_i(x_i)\\
N parameters & RV_i(x0_i,\dots,xn_i)     & RV0_i(x0_i,\dots,xn_i), RV1_i(x0_i,\dots,xn_i)    & RV0_i(x0_i,\dots,xn_i), \dots, RVn_i(x0_i,\dots,xn_i)\\
\end{tabular}
\caption{Each $vi$ is a distinct variable such as precipitation or temperature,
  each $xi$ is a distinct parameter such as time or space, and i is the
  observation identifier.}
\end{table}

The variables $v0,\dots, vn$ can also potentially be dependent on each other,
adding in a third level of interactions that a visualization may want to
caputure. While this survey focuses on data structured in a linear, tabular, or
data cube way, it could potentially be expanded to network data where the xi
parameters represent edges in the graph.

### 2016-11-03 sections/section1d.tex
moved section text into proper sections

\begin{document}
\section{1D Fields}

\subsection{Independent Variables}

\subsection{Univariate Observations}
%%functional box + derivatives,

>>>>

\begin{document}
\section{1D Fields}
While boxplots were designed to visualize discrete observations, often there
are no discrete observations and instead the `observation` is a timeseries
curve (where the connections are as important as the individual points) or
spatial region. Understanding the distributions of functional observations
often relies first ordering the functions such that there's some measure of how
close these functions are to each other--functional depth\cite{febrero2007},
bivariate score depth\cite{hyndman2009}, and
bivariate kernel density estimation\cite{scott1992} are some examples. This
ordering then yields functional analogs to median and other percentile
metrics. For small values, the ordered curves can then be colored based on the
ordering metric \cite{hyndman2009}, but this method is unweildy for larger
ensembles.

\subsection{Univariate Observations}

\begin{figure}
  %%\includegraphics{funcbaghdr}
  \label{fig:funcbag}
\end{figure}
Instead, Hyndmand and Shang proposed functional versions of the bag and HDR
plots\cite{hyndman2009}, as shown in Figure~\ref{fig:funcbag}. In these graphs,
what was traditionally the box part of a box plot is now a region bounded by
the curves (obtained by using one of the previously mentioned ordering
metrics). The median curve is defined as a blackline and the exceptional
outlines are shown. By using regions defined by the curves, the functional bag
and HDR plots display the variance in the ensembles while simulatanously
flattening the similaraties so that the outliers are strongly highlighted. The
HDR method better highlights outliers closer to the median, whereas the bagplot
method better highlights the outliers outside the envelopes of functions.

The functional boxplot was developed to describe the distributional nature of
functional data\cite{sun2011}. \textbf{expand this section with discusson of
  Sun's paper}

\begin{figure}
  \includegraphics[width=.75\linewidth]{contour_weather}
  \caption{\label{fig:countour}}
\end{figure}
While functional bag, box, and HDR plots all show outliers, each method is most
successful visualizing a specific type of outlier. Countour
boxplots\cite{whitaker2013} aim instead to capture the variablity of the
uncertainity via a band depth ordering metric. Each ensemble members band depth
is computed as sum of the probabilities that the observations in any given
ensemble fall within the max-min envelope defined by any two other
ensembles. The bands are then sorted by band depth such that the median is the
ensemble with about 50\% of its members withen all envelopes formed by other
bands (so most centered). Outer bands are chosen according to the task at
hand. In \cite{whitaker2013}, they apply the countour boxplots to temperature
visualizations. The authors argue that countour boxplots are an improved visual
idiom over the traditional spagehtti plots (%insert citation for them...)
because the spagehetti plots get nosy as the number of plots increase
and so it's hard to tease out specific patterns. Instead in
Figure~\ref{fig:countour}, the aut\
hors
remove most of the bands and instead visualize an outlier envelope(light gray)
and a more central envelope(dark gray). It retains much of the information of
the spagethhit, but removes the visual noise of the lines.
